\documentclass{article}
\usepackage[english]{babel}
\usepackage{xspace}
\usepackage{hyperref}

\newcommand\SI{SMTInterpol\xspace}
\newcommand\version{2.1-95-ge06bde5\xspace}

\title{\SI\\{\large Version \version}}

\author{J\"urgen Christ, Jochen Hoenicke\\
  University of Freiburg\\
  \texttt{\{christj,hoenicke\}@informatik.uni-freiburg.de}}

\begin{document}
\maketitle
\section*{Description}
\SI is a proof-producing SMT-solver written in Java.  It stores resolution
proof trees to compute interpolants~\cite{mcmillan05itp} used, e.~g., by
different model checking tools~\cite{HHP09,HHP10,DBLP:conf/vmcai/ErmisHP12}.

The solver reads input in SMTLIB format.  It includes parsers for DIMACS,
AIGER, and SMTLIB version 1.2 and 2.  All required and some optional commands
of the SMTLIB standard are supported.

All formulas are stored in a central term repository.  The repository
type-checks the formulas and does some simple boolean optimizations.  Asserted
formulas are converted to CNF using Plaisted--Greenbaum
encoding~\cite{DBLP:journals/jsc/PlaistedG86}.  The core of the solver is a
CDCL engine that is connected to multiple theories.  The engine uses these
theories during constraint propagation, backtracking, and consistency
checking.

For uninterpreted functions and predicates, we use a theory solver based on
the congruence closure algorithm~\cite{DBLP:conf/rta/NieuwenhuisO05}.  For
linear arithmetic, we use a theory solver based on the Simplex
algorithm~\cite{DBLP:conf/cav/DutertreM06}.  It always computes the strongest
bounds that can be derived for a variable and uses them during satisfiability
checks.  If a conflict cannot be explained using known literals, the solver
derives new literals and uses them in conflict explanation.  Disequalities are
resolved if they can be used to strengthen a bound.  Otherwise, they are
delayed until final checks.  The solver supports integer arithmetic using a
variant of the cuts from proof technique~\cite{DBLP:conf/cav/DilligDA09}
together with a branch-and-bound engine.  A new array decision procedure has
been added for this years competition.  It is based on the notion of
\emph{weak equivalences} and will be presented at the SMT workshop.

\SI uses a variant of model-based theory
combination~\cite{DBLP:journals/entcs/MouraB08}.  The linear arithmetic solver
does not propagate equalities between shared variables but introduces them as
decision points.  The model mutation algorithm resolves disequalities
and tries to create as many distinct equivalence classes as possible.

\section*{Competition Version}
The version submitted to the SMT-COMP 2014 is a preliminary release of version
2.2.  This release will include quantifier--free interpolation for the theory
of arrays which is still missing in the current solver.

The competition version can be downloaded from
\begin{center}
  \url{http://ultimate.informatik.uni-freiburg.de/smtinterpol/}
\end{center}
The sources are available via GitHub
\begin{center}
  \url{https://github.com/juergenchrist/smtinterpol}
\end{center}

Magic Number: 6649887

\bibliography{sysdec}
\bibliographystyle{alpha}
\end{document}
